\textbf{A framework for effective application of machine learning to microbiome-based classification problems}

### 2020-02-17 submission/manuscript.Rmd
added Jenna's comments

We highlighted the importance of model interpretation to gain greater biological insights into microbiota-associated diseases. In this study, we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance. Previous studies have emphasized the difficulty of interpreting the feature coefficients in linear models [@breiman_statistical_2001] and the biases introduced by computing feature importance using built-in methods (e.g., gini drop) of tree-based models [@strobl_bias_2007]. Therefore, we encourage our audience to use both interpretation methods highlighted in this study.  Human-associated microbial communities have complex correlation structures that create collinearity in the datasets. This can hinder our ability to reliably interpret models because the feature weights of correlated OTUs are influenced by one another [@dormann_collinearity:_2013]. To capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. When we used permutation importance, we partially accounted for collinearity by grouping correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other; however, we could reduce the correlation threshold to further investigate the relationships among correlated features. By our approach, we identified 432 OTUs out of 6,920 that had perfect correlations with at least one other OTU. The decision to establish correlation thresholds is left to researchers to implement for their own analyses. Regardless of the threshold, undestanding the correlation structures within the data is critical to avoid misinterpreting the models. Such structures are likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronounced due to many genes being correlated with one another because they come from the same chromosome. Finally, true causal mechanisms (e.g., role of microbiome in colorectal cancer) cannot be explained solely by the highest performing machine learning model [@li_accurate_2020]. To identify the true underlying microbial factors of a disease, it is crucial to follow up on any correlation analyses with further hypothesis testing and experimentation for biological validation.

@@

Our study highlights the need to make educated choices at every step of developing an ML model with microbiome data. We created an aspirational rubric that researchers can use to identify potential pitfalls when using ML in microbiome studies and ways to avoid them [Table S1]. We have highlighted the trade-offs between model complexity and interpretability, the need for tuning hyperparameters, the utility of held-out test sets for evaluating predictive performance, and the importance of considering correlation structures in datasets for reliable interpretation. Furthermore, we underscored the importance of proper experimental design and methods to help us achieve the level of validity and accountability we want from models built for patient health.

>>>>

We highlighted the importance of model interpretation to gain greater biological insights into microbiota-associated diseases. In this study, we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance. Previous studies have emphasized the difficulty of interpreting the feature coefficients in linear models [@breiman_statistical_2001] and the biases introduced by computing feature importance using built-in methods (e.g., gini drop) of tree-based models [@strobl_bias_2007]. Therefore, we encourage our audience to use both interpretation methods highlighted in this study as permutation importance is a model-agnostic tool that can be used to compared feature importance across different models.  Human-associated microbial communities have complex correlation structures that create collinearity in the datasets. This can hinder our ability to reliably interpret models because the feature weights of correlated OTUs are influenced by one another [@dormann_collinearity:_2013]. To capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. When we used permutation importance, we partially accounted for collinearity by grouping correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other; however, we could reduce the correlation threshold to further investigate the relationships among correlated features. By our approach, we identified 432 OTUs out of 6,920 that had perfect correlations with at least one other OTU. The decision to establish correlation thresholds is left to researchers to implement for their own analyses. Regardless of the threshold, undestanding the correlation structures within the data is critical to avoid misinterpreting the models. Such structures are likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronounced due to many genes being correlated with one another because they come from the same chromosome. Finally, true causal mechanisms (e.g., role of microbiome in colorectal cancer) cannot be explained solely by the highest performing machine learning model [@li_accurate_2020]. To identify the true underlying microbial factors of a disease, it is crucial to follow up on any correlation analyses with further hypothesis testing and experimentation for biological validation.

@@

Our study highlights the need to make educated choices at every step of developing an ML model with microbiome data. We created an aspirational rubric that researchers can use to identify potential pitfalls when using ML in microbiome studies and ways to avoid them [Table S1]. We highlighted the trade-offs between model complexity and interpretability, the need for tuning hyperparameters, the utility of held-out test sets for evaluating predictive performance, and the importance of considering correlation structures in datasets for reliable interpretation. We showed the importance of interpretability for generating hypotheses to identify causal, biological relationships and for identifying inconsistencies in model setup. Furthermore, we underscored the importance of proper experimental design and methods to help us achieve the level of validity and accountability we want from models built for patient health.

### 2020-02-17 submission/manuscript.tex
added Jenna's comments

We highlighted the importance of model interpretation to gain greater
biological insights into microbiota-associated diseases. In this study,
we showcased two different interpretation methods: ranking each OTU by
(i) their absolute weights in the trained models and (ii) their impact
on the predictive performance based on permutation importance. Previous
studies have emphasized the difficulty of interpreting the feature
coefficients in linear models (37) and the biases introduced by
computing feature importance using built-in methods (e.g., gini drop) of
tree-based models (38). Therefore, we encourage our audience to use both
interpretation methods highlighted in this study. Human-associated
microbial communities have complex correlation structures that create
collinearity in the datasets. This can hinder our ability to reliably
interpret models because the feature weights of correlated OTUs are
influenced by one another (39). To capture all important features, once
we identify highly ranked OTUs, we should review their relationships
with other OTUs. These relationships will help us generate new
hypotheses about the ecology of the disease and test them with follow-up
experiments. When we used permutation importance, we partially accounted
for collinearity by grouping correlated OTUs to determine their impact
as a group. We grouped OTUs that had a perfect correlation with each
other; however, we could reduce the correlation threshold to further
investigate the relationships among correlated features. By our
approach, we identified 432 OTUs out of 6,920 that had perfect
correlations with at least one other OTU. The decision to establish
correlation thresholds is left to researchers to implement for their own
analyses. Regardless of the threshold, undestanding the correlation
structures within the data is critical to avoid misinterpreting the
models. Such structures are likely to be a particular problem with
shotgun metagenomic datasets where collinearity will be more pronounced
due to many genes being correlated with one another because they come
from the same chromosome. Finally, true causal mechanisms (e.g., role of
microbiome in colorectal cancer) cannot be explained solely by the
highest performing machine learning model (40). To identify the true
underlying microbial factors of a disease, it is crucial to follow up on
any correlation analyses with further hypothesis testing and
experimentation for biological validation.

@@

Our study highlights the need to make educated choices at every step of
developing an ML model with microbiome data. We created an aspirational
rubric that researchers can use to identify potential pitfalls when
using ML in microbiome studies and ways to avoid them {[}Table S1{]}. We
have highlighted the trade-offs between model complexity and
interpretability, the need for tuning hyperparameters, the utility of
held-out test sets for evaluating predictive performance, and the
importance of considering correlation structures in datasets for
reliable interpretation. Furthermore, we underscored the importance of
proper experimental design and methods to help us achieve the level of
validity and accountability we want from models built for patient
health.

>>>>

We highlighted the importance of model interpretation to gain greater
biological insights into microbiota-associated diseases. In this study,
we showcased two different interpretation methods: ranking each OTU by
(i) their absolute weights in the trained models and (ii) their impact
on the predictive performance based on permutation importance. Previous
studies have emphasized the difficulty of interpreting the feature
coefficients in linear models (37) and the biases introduced by
computing feature importance using built-in methods (e.g., gini drop) of
tree-based models (38). Therefore, we encourage our audience to use both
interpretation methods highlighted in this study as permutation
importance is a model-agnostic tool that can be used to compared feature
importance across different models. Human-associated microbial
communities have complex correlation structures that create collinearity
in the datasets. This can hinder our ability to reliably interpret
models because the feature weights of correlated OTUs are influenced by
one another (39). To capture all important features, once we identify
highly ranked OTUs, we should review their relationships with other
OTUs. These relationships will help us generate new hypotheses about the
ecology of the disease and test them with follow-up experiments. When we
used permutation importance, we partially accounted for collinearity by
grouping correlated OTUs to determine their impact as a group. We
grouped OTUs that had a perfect correlation with each other; however, we
could reduce the correlation threshold to further investigate the
relationships among correlated features. By our approach, we identified
432 OTUs out of 6,920 that had perfect correlations with at least one
other OTU. The decision to establish correlation thresholds is left to
researchers to implement for their own analyses. Regardless of the
threshold, undestanding the correlation structures within the data is
critical to avoid misinterpreting the models. Such structures are likely
to be a particular problem with shotgun metagenomic datasets where
collinearity will be more pronounced due to many genes being correlated
with one another because they come from the same chromosome. Finally,
true causal mechanisms (e.g., role of microbiome in colorectal cancer)
cannot be explained solely by the highest performing machine learning
model (40). To identify the true underlying microbial factors of a
disease, it is crucial to follow up on any correlation analyses with
further hypothesis testing and experimentation for biological
validation.

@@

Our study highlights the need to make educated choices at every step of
developing an ML model with microbiome data. We created an aspirational
rubric that researchers can use to identify potential pitfalls when
using ML in microbiome studies and ways to avoid them {[}Table S1{]}. We
highlighted the trade-offs between model complexity and
interpretability, the need for tuning hyperparameters, the utility of
held-out test sets for evaluating predictive performance, and the
importance of considering correlation structures in datasets for
reliable interpretation. We showed the importance of interpretability
for generating hypotheses to identify causal, biological relationships
and for identifying inconsistencies in model setup. Furthermore, we
underscored the importance of proper experimental design and methods to
help us achieve the level of validity and accountability we want from
models built for patient health.
